A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-04-24 jacobian_calc_CMAME.tex
Fixed validation section reference, some spelling

\section{\todo[]{Re: I.5}Simple example of CS, AD and FD for solution of nonlinear problems.}
\label{sec:WorkingExamples}

In order to illustrate a comparison of the numerical differentiation techniques in a simpler form than it appears in the large-scale analysis code covered in the following sections, a simple example borrowed from \cite{rezaiee2010dynamic} was implemented with each of the methods, CS, AD, and FD. All the source code to reproduce the results described here are available in the associated paper repository. This source describes the implemenation and interface to the \emph{Trilinos Secado library} for automatic differentiation.

@@

For this system, the equalibrium reaction force associated with a displacement $D$ of the spring is shown in equation~(\ref{eqn:TrussForce}) 

@@

\todo[]{Re: I.2} Peridynamics \cite{silling2000ret,silling:psa,silling2010peridynamic} is a nonlocal reformulation of the partial-differential equations that provide the statement of momentum balance in classical continuum mechanics. Its primary goal is to avoid the use of spatial derivatives in the balance equations or constitutive laws such that discontinuous displacements, i.e. cracks, are mathematically consistent with the governing equations. It has demonstrated great promise in modeling problems with pervasive failure \cite{littlewood2010}, interesting phenomenon such as dynamic crack branching \cite{ha2010sod}, and is being extended to multiphyics phenomena \cite{bobaru2011peridynamic,katiyar2013} and the problem of multi-scale coupling to molecular simulations \cite{seleson2009peridynamics,seleson2014atom,rahman2013b}.  The material model used in this study was a linear peridynamic solid which replicates linear elasticity in the \emph{local limit}, i.e.\ when the non-local length scale vanishes. While \emph{linear} may appear in the name of the material model, it is geometrically non-linear under large deformations and requires a nonlinear solver to converge to equalibrium conditions in these situations.  The peridynamic equation of motion is shown in equation~(\ref{eqn:PDMotion})
%
\begin{equation} 
    \rho(\mathbf{x})\mathbf{\ddot{u}(x)} = \int_{\mathcal{H_{\delta(\mathbf{x})}}} \mathbf{f(x,x'}) {\rm d}V_{\mathbf{x'}} + \mathbf{b(x)},
    \label{eqn:PDMotion}
\end{equation}

@@

\subsubsection{\todo[]{Re: I.2}The tangent stiffness matrix in a quasi-static peridynamic simulation}
The tangent-stiffness matrix is the Fr\'{e}chet derivative of the residual with respect to the deformed positions of the discrete computational nodes.  For the purposes of this work, the residual represents the unbalanced forces in the peridynamic equalibrium equation.  For the finite difference type methods in \emph{Peridigm}, a perturbation method is employed by probing the deformed configuration one degree-of-freedom at a time and so allowing the derivative of the residual for all nodes to be computed.  The resulting spatial rate of change in the residual for each of these independent probes becomes one column of the tangent-stiffness matrix.

@@

\emph{Peridigm} was primarily written by computational scientists and engineers; therefore, many advanced techniques in C++ such as multiple virtual inheritance, pointer arithmetic, and distributed data structures are employed. These concepts are not vital to understanding the CS method itself, but may be necessary to understand CS in \emph{Peridigm}. To see CS and other methods in a more narrow context, it is advised that readers also take a look at the simple examples, such as that described in Section~\ref{sec:WorkingExamples},  provided on the corresponding author's website.

@@

The methods were compared by running two sets of test problems where each method would solve a problem concurrently. In this context, concurrently means successively, drawing from the same independent variables, yet within the same computational process; this definition is characterized in further detail in Section~\ref{JGAM}. Each test problem simulated a $\SI{4}{\meter}$ by $\SI{0.5}{\meter}$ by $\SI{0.5}{\meter}$ meter block of a material undergoing tension along the axis parallel to the long dimension of the block. The test problems were set up in \emph{Peridigm} as quasi-static equilibrium problems, where the displacement boundary conditions used to apply tension were applied gradually in ``load steps'', and an equilibrium solution for each load step was achieved before applying the next load step. Applying the load in gradual steps is a widely used technique to ensure the Newton's method is always initialized near the actual solution and therefore likely to converge.  The fictional material represented in the model had values of $\SI{1.515e4}{\mega\pascal}$ for bulk modulus, and $\SI{7.813e4}{\mega\pascal}$ for shear modulus. The peridynamic horizon, a length scale that effectively sets the bandwidth of tangent-stiffness matrix in the computation, used in the test problems was always set to three times the nominal discretization size (i.e.  node spacing in the particle discretization scheme). Pictured in Figure.~\ref{fig:pd_discrete} is an illustration of the peridynamic discretizations used in this study. The coloration in the images indicate material regions that were defined to assist in post-processing application of boundary conditions. 
%
\begin{figure}[tbp] 
    \centering 
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{./figs/mesh2000_replacement.png}
        \caption{2000 node mesh.}
        \label{fig:MCMesh}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{./figs/mesh1000000_replacement.png}
        \caption{1000000 node mesh.}
        \label{fig:SCMesh}
    \end{subfigure}
    \caption{Peridynamic discritizations used in study.}
    \label{fig:pd_discrete}
\end{figure}

>>>>

\section{\todo[]{Re: I.5}Simple example of CS, AD and FD for solution of nonlinear problems.}
\label{subsec:Validation}

In order to illustrate a comparison of the numerical differentiation techniques in a simpler form than it appears in the large-scale analysis code covered in the following sections, a simple example borrowed from \cite{rezaiee2010dynamic} was implemented with each of the methods, CS, AD, and FD. All the source code to reproduce the results described here are available in the associated paper repository. This source describes the implementation and interface to the \emph{Trilinos Secado library} for automatic differentiation.

@@

For this system, the equilibrium reaction force associated with a displacement $D$ of the spring is shown in equation~(\ref{eqn:TrussForce}) 

@@

\todo[]{Re: I.2} Peridynamics \cite{silling2000ret,silling:psa,silling2010peridynamic} is a nonlocal reformulation of the partial-differential equations that provide the statement of momentum balance in classical continuum mechanics. Its primary goal is to avoid the use of spatial derivatives in the balance equations or constitutive laws such that discontinuous displacements, i.e. cracks, are mathematically consistent with the governing equations. It has demonstrated great promise in modeling problems with pervasive failure \cite{littlewood2010}, interesting phenomenon such as dynamic crack branching \cite{ha2010sod}, and is being extended to multiphyics phenomena \cite{bobaru2011peridynamic,katiyar2013} and the problem of multi-scale coupling to molecular simulations \cite{seleson2009peridynamics,seleson2014atom,rahman2013b}.  The material model used in this study was a linear peridynamic solid which replicates linear elasticity in the \emph{local limit}, i.e.\ when the non-local length scale vanishes. While \emph{linear} may appear in the name of the material model, it is geometrically non-linear under large deformations and requires a nonlinear solver to converge to equilibrium conditions in these situations.  The peridynamic equation of motion is shown in equation~(\ref{eqn:PDMotion})
%
\begin{equation} 
    \rho(\mathbf{x})\mathbf{\ddot{u}(x)} = \int_{\mathcal{H_{\delta(\mathbf{x})}}} \mathbf{f(x,x'}) {\rm d}V_{\mathbf{x'}} + \mathbf{b(x)},
    \label{eqn:PDMotion}
\end{equation}

@@

\subsubsection{\todo[]{Re: I.2}The tangent stiffness matrix in a quasi-static peridynamic simulation}
The tangent-stiffness matrix is the Fr\'{e}chet derivative of the residual with respect to the deformed positions of the discrete computational nodes.  For the purposes of this work, the residual represents the unbalanced forces in the peridynamic equilibrium equation.  For the finite difference type methods in \emph{Peridigm}, a perturbation method is employed by probing the deformed configuration one degree-of-freedom at a time and so allowing the derivative of the residual for all nodes to be computed.  The resulting spatial rate of change in the residual for each of these independent probes becomes one column of the tangent-stiffness matrix.

@@

\emph{Peridigm} was primarily written by computational scientists and engineers; therefore, many advanced techniques in C++ such as multiple virtual inheritance, pointer arithmetic, and distributed data structures are employed. These concepts are not vital to understanding the CS method itself, but may be necessary to understand CS in \emph{Peridigm}. To see CS and other methods in a more narrow context, it is advised that readers also take a look at the simple examples, such as that described in Section~\ref{subsec:Validation},  provided on the corresponding author's website.

@@

The methods were compared by running two sets of test problems where each method would solve a problem concurrently. In this context, concurrently means successively, drawing from the same independent variables, yet within the same computational process; this definition is characterized in further detail in Section~\ref{JGAM}. Each test problem simulated a $\SI{4}{\meter}$ by $\SI{0.5}{\meter}$ by $\SI{0.5}{\meter}$ meter block of a material undergoing tension along the axis parallel to the long dimension of the block. The test problems were set up in \emph{Peridigm} as quasi-static equilibrium problems, where the displacement boundary conditions used to apply tension were applied gradually in ``load steps'', and an equilibrium solution for each load step was achieved before applying the next load step. Applying the load in gradual steps is a widely used technique to ensure the Newton's method is always initialized near the actual solution and therefore likely to converge.  The fictional material represented in the model had values of $\SI{1.515e4}{\mega\pascal}$ for bulk modulus, and $\SI{7.813e4}{\mega\pascal}$ for shear modulus. The peridynamic horizon, a length scale that effectively sets the bandwidth of tangent-stiffness matrix in the computation, used in the test problems was always set to three times the nominal discretization size (i.e.  node spacing in the particle discretization scheme). Pictured in Figure.~\ref{fig:pd_discrete} is an illustration of the peridynamic discretizations used in this study. The coloration in the images indicate material regions that were defined to assist in post-processing application of boundary conditions. 
%
\begin{figure}[tbp] 
    \centering 
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{./figs/mesh2000_replacement.png}
        \caption{2000 node mesh.}
        \label{fig:MCMesh}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{./figs/mesh1000000_replacement.png}
        \caption{1000000 node mesh.}
        \label{fig:SCMesh}
    \end{subfigure}
    \caption{Peridynamic discretizations used in study.}
    \label{fig:pd_discrete}
\end{figure}
